A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-01 tex/codatadp.tex
minor

\author {Andrew J. Reynolds\inst{1} \and Jasmin Christian Blanchette\inst{2,3}}
\authorrunning {A. J. Reynolds \and J. C. Blanchette}
\institute{
\'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
\and
Inria \& LORIA, Villers-l\`es-Nancy, France \\
Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, Germany
}

@@

For the purposes of this paper, a (co)datatype specification consists of
$l$~mutually recursive types which are either all datatypes or all codatatypes.
Each datatype $\delta$ is equipped by $m \ge 1$ constructors, and each
constructor for $\delta$ takes $n \ge 0$ arguments, each associated with a
selector, and returns a $\delta$ value. The argument types must be either
already existing types or be one of the newly introduced (co)datatypes. The
introduced names for the (co)datatypes, the constructors, and the selectors must
be distinct with already introduced ones and among themselves. Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, C_{11}\;(s_{111}{:}\; \sigma_{111}) \,\ldots\, (s_{11n_{11}}{:}\; \sigma_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\vdots} \\ % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\mid}}\; C_{1m}\;(s_{1m1}{:}\; \sigma_{1m1}) \,\ldots\, (s_{1mn_{1m}}{:}\; \sigma_{1mn_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{ij} : \sigma_{ij1}\times\cdots\times\sigma_{ijk_{ij}} \to \delta_i$
and $\const{s}_{ijk} : \delta_i \to \sigma_{ijk}$.

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall
outside this fragment. For example, the specification
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree} \,=\, \const{Node}\;\ty{int}\;\ty{forest} \\[-\jot]
      \keyw{and}\;\, \ty{forest} \,=\, \const{FNil} \mid \const{FCons}\; \ty{tree}\; \ty{forest}
\end{aligned}
\]
is supported, whereas
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{llist}(\alpha) & \,=\, \const{LNil} \mid \const{LCons}\; \alpha\; \ty{llist}(\alpha) \\
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}\; \ty{int}\; \ty{llist}(\ty{tree})
\end{aligned}
\]
is beyond the fragment considered here.
% (although it makes sense and is supported by Isabelle \cite{xxx}).

@@

@@

Our calculus consists of derivation rules that operate on a set $\Ec$ of equalities and disequalities between $\Sigma$-terms.
We will write $\tEc$ to denote the set of $\Sigma$-terms occurring in $\Ec$.
We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sigma$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $s^i_j( C_j( \vec t ) )$
are replaced by $t_i$, and moreover assume that all additional constraints added to $\Ec$ are normalized in this manner.

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Refl}
\(
\inferrule{
  t \in \tEc
}{
  \Ec := \Ec, t \teq t
}
\)
\qquad
\rn{Symm}
\(
\inferrule{
 t_1 \teq t_2 \in \Ec
}{
 \Ec := \Ec, t_2 \teq t_1
}
\)
\qquad
\rn{Conflict}
\(
\inferrule{
  t \teq u, t \tneq u \in \Ec
}{
  \bot
}
\)
\\[3.7ex]
\rn{Trans}
\(
\inferrule{
  t_1 \teq t_2, t_2 \teq t_3 \in \Ec
}{
  \Ec := \Ec, t_1 \teq t_3
}
\)
\qquad
\rn{Cong} 
\(
\inferrule{
  \vec t \teq \vec u \in \Ec \quad f( \vec t ), f( \vec u ) \in \tEc
}{
  \Ec := \Ec, f( \vec t ) \teq f( \vec u )
}
\)
\\[3.7ex]
\rn{Unify$_1$} 
\(
\inferrule{
  C_1( \vec t ) \teq C_1( \vec u ) \in \Ec
}{
  \Ec := \Ec, \vec t \teq \vec u
}
\)
\qquad
\rn{Unify$_2$} 
\(
\inferrule{
  C_1( \vec t ) \teq C_2( \vec u ) \in \Ec
}{
  \bot
}
\)
\end{tabular}
\caption{Rules for bidirectional closure.
}
\label{fig:cc-rules}
\end{figure}

@@

To explain the rules in Figure~\ref{fig:ab-rules}, we introduce the following notions.
First, an edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is a transition between two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
(which we will call its \emph{source} and \emph{destination} respectively),
labelled by a constructor $C$ and integer $n$.
We say that edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is \emph{induced by $\Ec$} if and only if $C( \vec u ) \in \ec{t_1}$ for some $\vec u$ and $\ec{t_2} = \ec{u_n}$.
A \emph{path} $p$ is a non-empty ordered list of edges $e_1, \ldots e_n$ such that the destination of $e_i$ is the source of $e_{i+1}$ for $i = 1, \ldots (n-1)$,
and the sources of $e_1, \ldots, e_n$ are distinct.
We say that path $p$ is \emph{induced by $\Ec$} if and only if each of its edges are induced by $\Ec$.
Notice that since $\Ec$ contains a finite set of equivalence classes, all paths induced by $\Ec$ have finite length.
We call the source of a path is the source of its first edge, and the destination of a path is the destination of its last edge.
The set $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ is the smallest set containing all paths with source $\ec{t_1}$ and destination $\ec{t_2}$ that are induced by $\Ec$.
The set $\tpath{\Ec}{\ec{t_1}}$ is the union of $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ for all $t_2 \in \tEc$.

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Split} 
\(
\inferrule{
  s^i_j( t ) \in \tEc, \text{ or } t : \tau \text{ finite }
}{
  \Ec := \Ec, \tester{C_1}( t ) \ror \ldots \ror \Ec := \Ec, \tester{C_n}( t ) 
}
\)
\end{tabular}
\caption{Rule for splitting.  We write $\tester{C_i}( t )$ as shorthand for the equality $t \teq C_i( s^1_i( t ), \ldots, s^n_i( t ) )$.
}
\label{fig:split-rule}
\end{figure}

>>>>

\author {Andrew J. Reynolds\inst{1} \and Jasmin Christian Blanchette\inst{2,3}}
\authorrunning {A. J. Reynolds \and J. C. Blanchette}
\institute{
\'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
\and
Inria Nancy \& LORIA, Villers-l\`es-Nancy, France
\and
Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, Germany
}

@@

For the purposes of this paper, a (co)datatype specification consists of
$l$~mutually recursive types which are either all datatypes or all codatatypes.
Each datatype $\delta$ is equipped by $m \ge 1$ constructors, and each
constructor for $\delta$ takes $n \ge 0$ arguments, each associated with a
selector, and returns a $\delta$ value. The argument types must be either
already existing types or be one of the newly introduced (co)datatypes. The
introduced names for the (co)datatypes, the constructors, and the selectors must
be distinct with already introduced ones and among themselves. Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{11}(\bigl[\const{s}_{111}{:}\bigr]\; \sigma_{111}, \,\ldots,\, \bigl[\const{s}_{11n_{11}}{:}\bigr]\; \sigma_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\vdots} \\ % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\mid}}\; \const{C}_{1m}(\bigl[\const{s}_{1m1}{:}\bigr]\; \sigma_{1m1}, \,\ldots,\, \big[\const{s}_{1mn_{1m}}{:}\big]\; \sigma_{1mn_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{ij} : \sigma_{ij1}\times\cdots\times\sigma_{ijk_{ij}} \to \delta_i$
and $\const{s}_{ijk} : \delta_i \to \sigma_{ijk}$. The selector names are often
omitted in our examples, in which case default names are generated.

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall
outside this fragment. For example, the specification
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{forest}) \\[-\jot]
      \keyw{and}\;\, \ty{forest} & \,=\, \const{FNil} \mid \const{FCons}(\ty{tree},\, \ty{forest})
\end{aligned}
\]
is supported, whereas
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{llist}(\alpha) & \,=\, \const{LNil} \mid \const{LCons}(\alpha,\, \ty{llist}(\alpha)) \\
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{llist}(\ty{tree}))
\end{aligned}
\]
is beyond the fragment considered here.
% (although it makes sense and is supported by Isabelle \cite{xxx}).

@@

  * Introduce the discriminators already here.

@@

Our calculus consists of derivation rules that operate on a set $\Ec$ of equalities and disequalities between $\Sigma$-terms.
We will write $\tEc$ to denote the set of $\Sigma$-terms occurring in $\Ec$.
We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sigma$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $s^i_j( \const{C}_j( \vec t ) )$
are replaced by $t_i$, and moreover assume that all additional constraints added to $\Ec$ are normalized in this manner.

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Refl}
\(
\inferrule{
  t \in \tEc
}{
  \Ec := \Ec, t \teq t
}
\)
\qquad
\rn{Symm}
\(
\inferrule{
 t_1 \teq t_2 \in \Ec
}{
 \Ec := \Ec, t_2 \teq t_1
}
\)
\qquad
\rn{Conflict}
\(
\inferrule{
  t \teq u, t \tneq u \in \Ec
}{
  \bot
}
\)
\\[3.7ex]
\rn{Trans}
\(
\inferrule{
  t_1 \teq t_2, t_2 \teq t_3 \in \Ec
}{
  \Ec := \Ec, t_1 \teq t_3
}
\)
\qquad
\rn{Cong} 
\(
\inferrule{
  \vec t \teq \vec u \in \Ec \quad f( \vec t ), f( \vec u ) \in \tEc
}{
  \Ec := \Ec, f( \vec t ) \teq f( \vec u )
}
\)
\\[3.7ex]
\rn{Unify$_1$} 
\(
\inferrule{
  \const{C}_1( \vec t ) \teq \const{C}_1( \vec u ) \in \Ec
}{
  \Ec := \Ec, \vec t \teq \vec u
}
\)
\qquad
\rn{Unify$_2$} 
\(
\inferrule{
  \const{C}_1( \vec t ) \teq \const{C}_2( \vec u ) \in \Ec
}{
  \bot
}
\)
\end{tabular}
\caption{Rules for bidirectional closure.
}
\label{fig:cc-rules}
\end{figure}

@@

To explain the rules in Figure~\ref{fig:ab-rules}, we introduce the following notions.
First, an edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is a transition between two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
(which we will call its \emph{source} and \emph{destination} respectively),
labeled by a constructor $C$ and integer $n$.
We say that edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is \emph{induced by $\Ec$} if and only if $C( \vec u ) \in \ec{t_1}$ for some $\vec u$ and $\ec{t_2} = \ec{u_n}$.
A \emph{path} $p$ is a non-empty ordered list of edges $e_1, \ldots e_n$ such that the destination of $e_i$ is the source of $e_{i+1}$ for $i = 1, \ldots (n-1)$,
and the sources of $e_1, \ldots, e_n$ are distinct.
We say that path $p$ is \emph{induced by $\Ec$} if and only if each of its edges are induced by $\Ec$.
Notice that since $\Ec$ contains a finite set of equivalence classes, all paths induced by $\Ec$ have finite length.
We call the source of a path is the source of its first edge, and the destination of a path is the destination of its last edge.
The set $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ is the smallest set containing all paths with source $\ec{t_1}$ and destination $\ec{t_2}$ that are induced by $\Ec$.
The set $\tpath{\Ec}{\ec{t_1}}$ is the union of $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ for all $t_2 \in \tEc$.

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Split} 
\(
\inferrule{
  s^i_j( t ) \in \tEc, \text{ or } t : \tau \text{ finite }
}{
  \Ec := \Ec, \tester{\const{C}_1}( t ) \ror \ldots \ror \Ec := \Ec, \tester{\const{C}_n}( t ) 
}
\)
\end{tabular}
\caption{Rule for splitting.  We write $\tester{\const{C}_i}( t )$ as shorthand for the equality $t \teq \const{C}_i( s^1_i( t ), \ldots, s^n_i( t ) )$.
}
\label{fig:split-rule}
\end{figure}
